---
title: "Airbnb Price Prediction"
subtitle: Mihai-Bogdan Bîndilă 3264424
output:
  html_document:
    df_print: paged
---

```{r import_packages, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(conflicted)
library(tidyverse)
library(rsample)
library(randomForestSRC)
library(caret)
```

### Problem description

The number of listings on Airbnb has recorded a [steady increase](https://www.businessofapps.com/data/airbnb-statistics/), from 300,000 in 2013 to 7.7 million in 2023. This project aims to leverage this rich data to identify key factors influencing nightly rental prices. With a focus on Amsterdam as a case study, we will build a predictive model to estimate these prices. The developed model will be integrated into a web application designed for two user groups: 

**Owners:** Gain insights into optimal pricing strategies based on property characteristics and potential guest ratings. \
**Guests:** Assess the fairness of listed prices and explore price variations across different locations for similar properties.

### Data description

It contains the following features that will be used to predict the price per night (realSum): \
**- property characteristics:** type of room, private or shared, capacity, number of bedrooms, for business purposes or not \
**- data from Tripadvisor:** accessibility to attractions and restaurants within the neighborhood of the property \
**- location:** longitude, latitude, distance from the city center and from the nearest metro station \
**- ratings:** cleanliness and guest satisfaction scores \
**- host:** is super host or not


```{r load_data}
data_weekdays <- read_delim(
  file = "./data/amsterdam_weekdays.csv",
  delim = ",",
  col_names = TRUE,
  show_col_types = FALSE
)
data_weekends <- read_delim(
  file = "./data/amsterdam_weekends.csv",
  delim = ",",
  col_names = TRUE,
  show_col_types = FALSE
)
complete_data <- bind_rows(data_weekdays, data_weekends)
complete_data
```


### Data preprocessing

```{r factor_categorical_features}
complete_data <- complete_data %>%
  mutate(room_type = factor(room_type)) %>%
  mutate(room_shared = factor(room_shared)) %>%
  mutate(room_private = factor(room_private)) %>%
  mutate(host_is_superhost = factor(host_is_superhost)) %>%
  mutate(multi = factor(multi)) %>%
  mutate(biz = factor(biz))
```

### Data cleaning
- drop features w/o variance or unimportant
- check is there are missing values

```{r data_cleaning}
data <- subset(complete_data, select = -c(...1, attr_index, rest_index))
data

data <- rename(
  data,
  price = realSum,
  business_room = biz,
  is_superhost = host_is_superhost,
  multiple_rooms = multi,
  satisfaction_rating = guest_satisfaction_overall,
  center_dist = dist,
  attractions_index = attr_index_norm,
  restaurants_index = rest_index_norm,
  longitude = lng,
  latitude = lat,
  capacity = person_capacity
)


data$multiple_rooms <- ifelse(data$multiple_rooms == 0, FALSE, TRUE)
data$business_room <- ifelse(data$business_room == 0, FALSE, TRUE)

data$cleanliness_rating  <- data$cleanliness_rating * 10

data
```

### Data visualization
- check if there are outliers
```{r data_visualization}

data
```

### Feature engineering
- feature selection

### Modeling
- train
- optimize
- validate
- interpret weights / feature importance
- perform statistical tests

```{r data_splitting, warning=FALSE}
set.seed(19)

dataset_splitter <- initial_split(data, prop = 0.8)
train_data <- as.data.frame(training(dataset_splitter))
test_data <- as.data.frame(testing(dataset_splitter))
```

```{r modeling}
rf_model <- rfsrc(price ~ ., data = as.data.frame(data), importance = TRUE, )
rf_model

predictions <- predict(rf_model, newdata = test_data)
rsq <- function (x, y)
  cor(x, y) ^ 2
rf_model

rsq(predictions$predicted, test_data$price)
sqrt(mean((
  test_data$price - predictions$predicted
) ^ 2))

control <- trainControl(method = "cv", number = 3)
rf_model <- train(price ~ .,
                  data = data,
                  method = "rf",
                  trControl = control)
rf_model
```

### Final model

```{r final_model}
rf_final_model <- rfsrc(price ~ ., data = as.data.frame(data), importance = TRUE)
rf_final_model

feature_importance <- as.data.frame(rf_final_model$importance) %>%
  rename("importance" = "rf_final_model$importance")
feature_importance$importance  <- rf_final_model$importance / sum(feature_importance$importance) * 100
feature_importance <-  feature_importance %>% arrange(desc(importance))

ggplot(feature_importance, aes(x  = rownames(feature_importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#FF5A5F") +
  labs(title = "Scaled Feature Importance", x = "Feature", y = "Importance") +
  theme_minimal() +
  scale_y_continuous(breaks = seq(0, 100, by = 5),
                     labels = seq(0, 100, by = 5),) +
  scale_x_discrete(limits = rownames(feature_importance)) +
  coord_flip()
```

### Reflection
